---
title: "Installing Missing Packages from Bioconductor, CRAN and Github" 
author: "Ernest Omane-Kodie"
---

Over the past few days I have had to use computers that are different from my trusty old laptop. One thing I noticed while working with R on the new machines is that the excitement in checking for, and installing, missing packages wears off   pretty quickly. This blog post is a walk-through of a function I wrote to make the process less painful.


## A bit more background
When R developers create open source packages, they typically distribute them through [The Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/), [Github](https://github.com/) or, if the package concerns computational biology and/or bioinformatics, [Bioconductor.](https://bioconductor.org/) 

This adds to the complication of knowing which source to install a package from, which installation functions to use and, in the case of GitHub packages, the exact repository path to use.

### Here are a few examples for context.

The `taskscheduleR` package is distributed on [Github.](https://github.com/) We can install it with:
```{r eval=FALSE, include=TRUE}
devtools::install_github("bnosac/taskscheduleR")
```

The `GenomicFeatures` package is distributed on [Bioconductor](https://www.bioconductor.org/install/) so to install it we do this:
```{r eval=FALSE, include=TRUE}
source("https://bioconductor.org/biocLite.R")
biocLite("GenomicFeatures")
```

Lastly, `Rcpp` is distributed on [CRAN](https://cran.r-project.org/) so we install it with:
```{r eval=FALSE, include=TRUE}
install.packages("Rcpp")
```

## The solution
We will use the `remedy` package as an example to test our code. To install `remedy`, we first need to figure out which repository it is distributed on:
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(glue)
package = "remedy" 
```

Is it available on CRAN?

We can use `tools::CRAN_package_db()` to extract metadata for the current packages in the CRAN package repository:
```{r}
package %in% tools::CRAN_package_db()$Package
```
As you can see, `remedy` is not on CRAN.

Is it on Biconductor? 

`BiocInstaller::all_group()` gives the names of all current packages on Bioconductor: 
```{r}
package %in% BiocInstaller::all_group()
```
`remedy` is not available on Bioconductor either. 

Lastly, we check whether the package is on GitHub. For this, we will use the [rpkg API](http://rpkg-api.gepuro.net/) and the `jsonlite` package. 
```{r}
url <- glue::glue("http://rpkg-api.gepuro.net/rpkg?q={package}")
gh_pkgs <- jsonlite::fromJSON(url)
gh_pkgs
```
Note that `fromJSON` extracts GitHub R package information from the `rpkg API` into a dataframe. If the said package is not available on Github, the funciton returns an empty list.

In our example, we can see that the package we're after is available on GitHub.
We will go ahead and install it in a while, but in the meantime let's tidy up the GitHub package metadata we just extracted:

```{r eval=TRUE, include=TRUE}
gh_pkg <- gh_pkgs %>% 
  separate(col = pkg_name, into = c("repo", "pkg"), sep = "/", remove = F) %>% 
  filter(pkg == package) %>% 
  select(pkg_name, repo, pkg)
gh_pkg
```

The package is in the ThinkR-open repository on GitHub so we can install it with:
```{r eval=FALSE, include=TRUE}
install_github(gh_pkg$pkg_name[1])
```

We can follow a similar process to install packages on CRAN and Bioconductor.


## Putting it all together
Let's finish off by wrapping everything in a function and throwing in a few useful checks for robustness:
```{r eval=FALSE, include=TRUE}
install_missing_pkg <- function(package){
  url <- glue::glue("http://rpkg-api.gepuro.net/rpkg?q={package}")
  cran_pkgs <- tools::CRAN_package_db()$Package
  gh_pkgs <- jsonlite::fromJSON(url)
  source("http://bioconductor.org/biocLite.R")
  bioc_pkgs = all_group()
  
  if (is.null(nrow(gh_pkgs)) | !(package %in% union(cran_pkgs, bioc_pkgs))) {
    stop("`package` is not available on CRAN, Github or Bioconductor")
  }
  # install from CRAN  
  if (package %in% cran_pkgs){
    install.packages(package)
  # install from Bioconductor    
  } else if (package %in% bioc_pkgs){
    biocLite(package, suppressUpdates = TRUE)
  }
  # install from Github
  else{
    gh_pkg <- gh_pkgs %>% 
      separate(col = pkg_name, into = c("repo", "pkg"), sep = "/", remove = F) %>% 
      mutate(available = package %in% cran_pkgs) %>% 
      filter(pkg == package)
    devtools::install_github(gh_pkg$pkg_name[1])
  }
}
```


The full blog post is available on [www.eokodie.com.](https://www.eokodie.com/blog/installing-missing-packages-from-bioconductor-cran-and-github/)